Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-15 ms.tex
revised migration model some more

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and fitness-associated outcrossing,
and exhibits the paraphrase: \emph{"Necessity is the mother of genetic invention."}
\end{abstract}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (I-C+MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

In terms of the general model, we can write
\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M_{i,j} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix}.
\end{equation}
Here, too, $M$ and $(I-C+MC)D$ are irreducible.

% Migration model
\paragraph*{Migration model.}\label{sec:mig_model_1}
% TODO explain why sum x y is 1 and wbar is sum xw yq
Here we assume a large haploid population of haploids that occupies two demes (the model can be easily extended to an arbitrary number demes).
Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, and $\sum_{k=1}^{n}{x_k + y_k} = 1$.

In this model, the fitness of the individual is determines by both its type and its deme, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes.
Therefore, the fitness of genotypes $A_k$ is $w_{1,k}$ in deme 1 and $w_{2,k}$ in deme 2 for $k=1,2,\ldots,n$, and the population mean fitnesses of the entire population is $\bar{w} = \sum_{k \ge 0}{w_{1,k} x_k + w_{2,k} y_k}$.
In addition, we assume that the fitness of type $k$ is different in deme 1 and 2, that is, $w_{1,k} \ne w_{2,k}$ for any $k=1,\ldots,n$.

The probability that offspring of an individual from deme $a$ with type $A_k$ return to their parent home deme is $1-m_{j,k}$.
That is, migration rates also depend on both the type and the deme.

Following selection and migration the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_{1,k}) \frac{w_{1,k}}{\bar{w}} x_k + m_{2,k} \frac{w_{2,k}}{\bar{w}} y_k \\
y_k' = m_{1,k} \frac{w_{1,k}}{\bar{w}} x_k + (1 - m_{2,k}) \frac{w_{2,k}}{\bar{w}} y_k
\end{align}
for $k=1,2,\ldots,n$.

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_{1,1}, \ldots, m_{1,n}, m_{2,1}, \ldots, m_{2,n}], \\
D = diag[w_{1,1}, \ldots, w_{1,n}, w_{2,1}, \ldots, w_{2,n}], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.
In this case, $M$ is block matrix with each block an irreducible 2-by-2 matrix, and so is $(I-C+MC)D$.

@@

This research was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.
}

>>>>

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis and condition-dependent dispersal,
and exhibits the paraphrase: \emph{"Necessity is the mother of genetic invention."}
\end{abstract}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (I-C+MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Here we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

In terms of the general model, we can write
\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix}.
\end{equation}
Here, too, $M$ and $(I-C+MC)D$ are irreducible.

% Migration model 1
\paragraph*{Migration model 1.}\label{sec:mig_model_1}

Here we assume a large haploid population of haploids that occupies $n$ demes, $A_1, \ldots, A_n$. In this case we directly use the notation of Eq.~\ref{eq:model}.
Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
The population mean fitness is $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$.

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{k,j}$, where $C_k \ge 0$, $M_{k,j} \ge 0$ and $\sum_{k=1}^{n}{M_{k,j}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $x'_k$ are given by Eq.~\ref{eq:model_sum}.
If the rows of $M$ are equal, that is, $M_{k,j}=m_k$ for all $j$,
then $m_k$ can be considered the relative population size of deme $A_k$;
this is the non-homogeneous extension to Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.
The transition matrix $M$ can account for other factors that determine the probability of non-homing juveniles from deme $A_j$ to migrate to deme $A_k$ -- for example, deme adjacency:
if demes are arranged in a circle, 
then $M$ can take the form
\begin{equation}
M = \begin{bmatrix}
p_0 & p_1 & \ldots & p_n \\
p_n & p_0 & \ldots & p_{n-1} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_n & \ldots & p_0
\end{bmatrix}.
\end{equation}
where $p_{-k} = p_{n-k}$, $p_k \ge 0$, and $\sum_{k=1}^{n}{p_k}=1$.

@@

This work was supported in part by 
the Department of Information and Computer Sciences at the University of Hawai`i at M{\=a}noa, 
the Konrad Lorenz Institute for Evolution and Cognition Research, 
the Mathematical Biosciences Institute through National Science Foundation Award \#DMS 0931642, 
the Stanford Center for Computational, Evolutionary and Human Genomics, 
and the Morrison Institute for Population and Resources Studies, Stanford University.
}
